### 2020-11-23 written/V1_model/V1_model.tex
checkpoint V1 model

\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{wrapfig}
\usepackage{fancyhdr}
\usepackage{tikz-qtree}
\usepackage{tikz-qtree-compat}
\usepackage[normalem]{ulem}
\usepackage{tikz}
\usepackage{graphicx}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator*{\argmax}{argmax}

@@

\section{EPI clarifies the implications of contrast-response flipping on [noise quenching, inhibition stabilization] in V1}\label{results_V1}
Dynamical models of excitatory (E) and inhibitory (I) populations with supralinear input-output function have succeeded in explaining a host of experimentally documented phenomena.
In a regime characterized by inhibitory stabilization of strong recurrent excitation, these models give rise to paradoxical responses \cite{tsodyks1997paradoxical}, selective amplification  \cite{murphy2009balanced}, surround suppression \cite{ozeki2009inhibitory} and normalization \cite{rubin2015stabilized}. 
Despite their strong predictive power, E-I circuit models rely on the assumption that inhibition can be studied as an indivisible unit. 
However, experimental evidence shows that inhibition is composed of distinct elements -- parvalbumin (P), somatostatin (S), VIP (V) --
composing 80\% of GABAergic interneurons in V1 \cite{markram2004interneurons, rudy2011three, tremblay2016}, and that these inhibitory cell types follow specific connectivity patterns (Fig. \ref{fig:SV_flip}A) \cite{pfeffer2013inhibition}.
Recent theoretical advances \cite{litwin2016inhibitory, GarciaDelMolino2017, Chen2019},  have only started to address the consequences of this multiplicity in the dynamics of V1, strongly relying on linear theoretical tools. 
Here, we use EPI to analyze the posteriors of a stochastic nonlinear dynamical model of V1 conditioned on the emergent property of contrast-dependent S-V flipping.
We then [use this info to make a statement about ...]

Specifically, we consider a four-dimensional circuit model with dynamical state given by the firing rate $x$ of each neuron-type population $x = \left[x_E, x_P , x_S, x_V \right]^\top$. Given a time constant of $\tau = 20$ ms and a power $n = 2$, the dynamics are driven by the rectified and sum of recurrent inputs $Wx$, external inputs $h$, and external noise $\epsilon \sim \mathcal{N}(0, \sigma_{\epsilon}^2)$:

\begin{equation}
\tau \frac{dx}{dt} = -x + [W x+ h + \epsilon]_+^n.
\end{equation}

We considered fixed effective connectivity weights $W$ approximated from experimental recordings of publicly available datasets of mouse V1 \cite{allen2018layer, billeh2019systematic} (see Section \ref{methods_V1}).
The input $h = b + dh$ is comprised of a baseline input  $b = \left[ b_E, b_P , b_S , b_V \right]^\top$ and a differential input $dh = \left[ dh_E , dh_P , dh_S , dh_V\right]^\top$ to each neuron-type population.  
Throughout subsequent analyses, the baseline input is $b = \left[ 1 ,1,1,1\right]^\top$.

With this model, we are interested in the differential responses of each neuron-type population to changes in input $dh$. 
Initially, we studied the linearized response of the system to input $\frac{dx_{ss}}{dh}$ at the steady state response $x_{ss}$, i.e. a fixed point. 
All analyses of this model consider the steady state response, so we drop the notation $ss$ from here on.
While this linearization accurately predicts differential responses $dx = \left[ dx_{E} , dx_{P} , dx_{S} ,dx_{V} \right]^\top$  for small differential inputs to each population $dh = \left[ 0.1 , 0.1 , 0.1 , 0.1 \right]^\top$ (Fig \ref{fig:V1_EPI}B left), the linearization is a poor predictor in this nonlinear model more generally (Fig. \ref{fig:V1_EPI}B right).  
Currently available approaches to deriving the steady state response of the system are limited.

\clearpage
\begin{figure}[h]
\caption{\small \textbf{A}.  Four-population model of primary visual cortex with excitatory (black), parvalbumin (blue), somatostatin (red), and VIP (green) neurons.   Some neuron-types largely do not form synaptic projections to others  (excitatory and inhibitory projections filled and unfilled, respectively).
\textbf{B}. EPI posterior $q_\theta(z \mid \mathcal{B}_{S-V})$ for S-V flipping. The obtained posterior is visualized as 500 samples from the inferred distribution colored by $\log(q_\theta(z))$. This posterior is bimodal and concentrated in planes $dh_E > 0$ and $dh_E < 0$ at distinct modes $z_1$ (black star) and $z_2$ (gray, star), respectively.  Bottom-left: Posterior predictive distribution of the emergent property statistics with respect to the constrained means (black, dashed line) and variances (gray, dashed lines at two stds). 
\textbf{C}. Posterior predictive distribution of $d_{S-V}(h)$ of each mode shows that the $z_1$-mode produces V-to-S flipping with increasing contrast and the $z_2$-mode produces S-to-V flipping.
\textbf{D}. Model simulations at the mode $z_1$ at $b_1$ (left), $b_1 + dh_1^{(c)}$ (middle), and steady state solutions for varying levels of contrast (right).  Shaded area is one standard deviation according to randomness of $\epsilon$.  \textbf{E}. Same as D. for $z_2$. \textbf{F}. ...
 }\label{fig:SV_flip}
\begin{center}
\includegraphics[scale=.31]{figs/Fig3/Fig3.pdf}
\end{center}
\end{figure}
\clearpage

TODO Summarize S-V flipping phenomena and why it's a good EP
TODO Explain that we can't really fit the whole data-set of responses yet because the model is quite constrained.
TODO Motivate this input model, where $dh^{(c)}$ is the direction of increasing contrast.

\begin{equation}
 \begin{bmatrix} h_E \\ h_P \\ h_S \\ h_V \end{bmatrix} =  \begin{bmatrix} b_E \\ b_P \\ b_S \\ b_V \end{bmatrix} +  \begin{bmatrix} dh^{(c)}_E \\ dh^{(c)}_P \\ 0 \\ 0 \end{bmatrix} 
\end{equation}

Since we are unsure what either the baseline input $b$ and contrast-dependent change in input $dh^{(c)}$ should be, we treat them as free parameters when running EPI.
\begin{equation}
 z =  \begin{bmatrix} b_E & b_P & b_S & b_V & dh^{(c)}_E & dh^{(c)}_P \end{bmatrix}^\top
\end{equation}
We consider positive baseline inputs $b_\alpha \in \left[0, 3 \right]$ and small contrast-dependent changes in input $|dh_\alpha| \leq 1$.  
To find parameters resulting in S-V flipping, we focus on models driven by nominal amounts of external noise $\sigma_{\eta} = 0.1$.

We formulate the emergent property of S-V flipping with a statistic measuring he difference in steady state ($x_ss$) between the S- and V-population at a given input:
\begin{equation}
d_{S-V}(h) = x_{ss,S}(h) - x_{ss,V}(h).
\end{equation}
For S and V to flip their steady states, the difference between the S- and V-population rates at $h = b$ must have opposite sign from $h = b + dh$.  Therefore, we stipulate the emergent property of S-V flipping to require the product between $d_{S-V}(b)$ and $d_{S-V}(b+dh)$ to be appreciably negative.  Second we stipulate that the differences between S and V in each input condition cancel out on average ($d_{S-V}(b) + d_{S-V}(b+dh)$ is 0 on average).  The means and variances of the emergent property statistics were sensibly chosen based on some inexpensive model simulations within the parameter bounds.

\begin{equation}
\mathcal{B}_{S-V} \triangleq \mathbb{E}\begin{bmatrix} d_{S-V}(b)d_{S-V}(b+dh) \\ \left( d_{S-V}(b)d_{S-V}(b+dh) - (-0.25) \right)^2 \\  d_{S-V}(b) + d_{S-V}(b+dh) \\  \left( d_{S-V}(b) + d_{S-V}(b+dh)\right)^2 \end{bmatrix} = 
\begin{bmatrix} -0.25 \\ 0.125^2 \\ 0 \\ 0.125^2 \end{bmatrix}
\end{equation}

We ran EPI to inspect the structure of the posterior distribution of $z$ conditioned on S-V flipping shown in Fig 2.  
It is clear from this visualization that S-V flipping in this V1-model is sensitive with respect to some parameter settings and robust with respect to others.
Additionally, the posterior is bimodal: there is one mode in each of the hyperplanes $dh_E < 0$ and $dh_E > 0$.

The structure of the V1 model S-V flipping posterior yields the following insights:
\begin{enumerate}
\item The marginal distributions of $q_\theta(h_P \mid \mathcal{B}_{S-V})$ and  $q_\theta(dh_P \mid \mathcal{B}_{S-V})$ are approximately uniform along their allowed range.  
The approximate uniformity of the $P-population$ parameter marginal distributions shows that the P-population plays little role in S-V flipping. 
\item The pairwise marginal distribution of $q_\theta(h_S, h_V \mid \mathcal{B}_{S-V})$ shows strong correlation between these parameters.
The strong correlation between $h_S$ and $h_V$ in the posterior reveals that S-V flipping is sensitive with respect to the baseline inputs to the S- and V-populations.
\textbf{Augment this to show that multidim info from Hessian is useful.}
The Hessian provided by EPI at the modes indicates that coordinated increases of $h_S$ and $h_V$ by a ratio of $\frac{h_S}{h_V} = 0.701$ will preserve S-V flipping, while changes in an orthogonal dimension will disrupt it.

\item This distribution is multimodal.
\end{enumerate}

@@

@@

\begin{figure}[h]
\caption{\small \textbf{A}. ...
 }\label{fig:drdh}
\begin{center}
\includegraphics[scale=.3]{figs/FigSabc/FigSabc.pdf}
\end{center}
\end{figure}

\begin{figure}[h]
\caption{\small \textbf{A}. ...
 }\label{fig:drdh}
\begin{center}
\includegraphics[scale=.12]{figs/FigSX/FigSX.pdf}
\end{center}
\end{figure}

@@

>>>>

\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{wrapfig}
\usepackage{fancyhdr}
\usepackage{tikz-qtree}
\usepackage{tikz-qtree-compat}
\usepackage[normalem]{ulem}
\usepackage{tikz}
\usepackage{systeme}
\usepackage{graphicx}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator*{\argmax}{argmax}

@@

\section{Exploratory analysis of V1 with EPI produces a novel theory}\label{results_V1}
Dynamical models of excitatory (E) and inhibitory (I) populations with supralinear input-output function have succeeded in explaining a host of experimentally documented phenomena.
In a regime characterized by inhibitory stabilization of strong recurrent excitation, these models give rise to paradoxical responses \cite{tsodyks1997paradoxical}, selective amplification  \cite{goldman2009memory, murphy2009balanced}, surround suppression \cite{ozeki2009inhibitory} and normalization \cite{rubin2015stabilized}. 
Despite their strong predictive power, E-I circuit models rely on the assumption that inhibition can be studied as an indivisible unit. 
However, experimental evidence shows that inhibition is composed of distinct elements -- parvalbumin (P), somatostatin (S), VIP (V) --
composing 80\% of GABAergic interneurons in V1 \cite{markram2004interneurons, rudy2011three, tremblay2016}, and that these inhibitory cell types follow specific connectivity patterns (Fig. \ref{fig:SV_flip}A) \cite{pfeffer2013inhibition}.
Recent theoretical advances \cite{litwin2016inhibitory, GarciaDelMolino2017, Chen2019},  have only started to address the consequences of this multiplicity in the dynamics of V1, strongly relying on linear theoretical tools. 
Here, we use EPI to gain a comprehensive understanding of how external circuit inputs govern S-V flipping and modulate variability.

We consider contrast responses of a nonlinear dynamical V1 circuit model (Fig. \ref{fig:SV_flip}A) with a state comprised of each neuron-type population's average membrane potential $x = \left[x_E, x_P , x_S, x_V \right]^\top$ (mV).
Specifically, we examine the stochastic stabilized supralinear network \cite{hennequin2018dynamical}, generalized to have inhibitory multiplicity (see Section \ref{methods_V1}).
Each population of neuron-type $\alpha$ receives recurrent input $\left(W f_r(x)\right)_\alpha$ from synaptic projections, where $W$ is the connectivity matrix and $f_r = \left[\cdot\right]_+^2$ is the rate nonlinearity.  
$W$ is calculated from recent experiments measuring post-synaptic potentials and connectivity rates in mice  \cite{allen2018layer, billeh2019systematic}.
Additionally, each population experiences a positive baseline input of $b_\alpha$, a contrast ($c$)-dependent input $h_\alpha$, and a stochastic input $\epsilon_\alpha$.
%Since contrast is only considered to change input to the E- and P-populations in V1 (\textbf{TODO} AP to add citation), $h_E = h_P = c h^{(c)}$ and $h_S = h_V = 0\text{mV}$.
Little is known about the scales of these baseline and contrast-dependent inputs, so we treat these as free parameters in our application of EPI:
 \begin{equation}
 z =  \begin{bmatrix} b_E & b_P & b_S & b_V & h^{(c)} \end{bmatrix}^\top.
\end{equation}

From first principles, we may attempt to fit or infer $z$ based on recorded V1 population responses at varying contrasts (\textbf{TODO} AP to add citation).
However, such a class of constrained models rarely produce good fits that generalize well.
Instead, we can use EPI to condition $z$ on one of the most salient properties of these data-sets called ``S-V flipping" --  a winner-take-all relationship between the S- and V-populations. At low contrast values, the V-population dominates the S-population, and at greater contrast values, the S-population dominates the V-population.  We formulate the emergent property of S-V flipping with a normalized statistic measuring the difference in steady state ($x_{ss}$) between the S- and V-population at a given contrast:
\begin{equation}
d_{S,V}(c; z) = \frac{f_r(x_{ss,S}(c; z)) - f_r(x_{ss,V}(c; z))}{|f_r(x_{ss}(c; z))|_2}
\end{equation}

\clearpage
\begin{figure}[h]
\caption{\small \textbf{A}.  Four-population model of primary visual cortex with excitatory (black), parvalbumin (blue), somatostatin (red), and VIP (green) neurons.   Some neuron-types largely do not form synaptic projections to others  (excitatory and inhibitory projections filled and unfilled, respectively).
\textbf{B}. EPI posterior $q_\theta(z \mid \mathcal{B}_{S-V})$ for S-V flipping. The obtained posterior is visualized as 500 samples from the inferred distribution colored by $\log(q_\theta(z))$. This posterior is bimodal and concentrated in planes $h^{(c)} > 0$ and $h^{(c)} < 0$ at distinct modes $z_1$ (black star) and $z_2$ (gray, star), respectively.  Bottom-left: Posterior predictive distribution of the emergent property statistics with respect to the constrained means (black, dashed line) and variances (gray, dashed lines at two standard deviations). 
\textbf{C}. Posterior predictive distribution of $d_{S,V}$ for each mode.  The $z_1$-mode produces V-to-S flipping with increasing contrast.
\textbf{D}. Model simulations at the mode $z_1$ at $c=0$ (solid) and $c=1$ (dashed). \textbf{E}. Mean system response at varying contrasts for $z_1$. Error bars show noise in rate. \textbf{F}. Plot of noise with contrast from E. Error bars are standard error measured across simulations.
 }\label{fig:SV_flip}
\begin{center}
\includegraphics[scale=.31]{figs/Fig3/Fig3.pdf}
\end{center}
\end{figure}
\clearpage

For S and V to flip their steady states, the difference between the S- and V-population rates at no contrast ($c=0$) must have opposite sign at full contrast ($c=1$).  Therefore, we stipulate the emergent property of S-V flipping to require the product of $d_{S,V}(c=0)$ and $d_{S,V}(c=1)$ to be appreciably negative.  The means and variance of this emergent property statistics were chosen such that the posterior predictive distribution always produced S-V flipping.

\begin{equation}
\begin{split}
\mathcal{B}_{S,V} \triangleq \text{\hspace{.5cm}} \mathbb{E}\left[ d_{S,V}(0)d_{S,V}(1) \right] &= -0.25 \\ 
\text{Var}\left[ d_{S,V}(0)d_{S,V}(1)\right] &=  0.125^2
\end{split}
\end{equation}

We ran EPI to obtain the parameter distribution of neuron-type population input parameters $z$ that produce S-V flipping (Fig. \ref{fig:SV_flip}B).
We considered positive baseline inputs $b_\alpha$ from 0 to 25mV and contrast-dependent changes in $h^{(c)}$ from -10 to 10mV.
The shape and structure of this distribution reveal some important properties of this model.
\begin{enumerate}
\item Two types of $z$ result in S-V flipping. (There are two modes separated by the hyperplane $h^{(c)} = 0$.)
\item S-V flipping is sensitive to anticorrelated changes in $b_S$ and $b_V$.  (The pairwise marginal distribution of $q_\theta(b_S, b_V \mid \mathcal{B}_{S,V})$ reveals strong correlation between these parameters.)
\item S-V flipping is most degenerate with respect to input to the P population. (The marginal distributions of $q_\theta(h_P \mid \mathcal{B}_{S,V})$ and  $q_\theta(dh_P \mid \mathcal{B}_{S,V})$ are approximately uniform along their allowed range.)
\end{enumerate}
Simulation from the posterior reveals that the biologically plausible mode ($h^{(c)} > 0$) matches experiments by producing V-to-S flipping with contrast.
Alternatively, the biologically implausible mode ($h^{(c)} > 0$) produces S-to-V flipping with contrast -- the opposite effect from experiments (Fig. \ref{fig:SV_flip}C).

We examined the responses of this circuit model to varying contrasts using the highest log-probability parameterization ($z_1$) of the plausible mode.
Simulations show the stochastic responses of the S- and V-population are indeed flipped at no and full contrast (Fig. \ref{fig:SV_flip}D).
This flipping occurs gradually with graded increases in contrast (Fig. \ref{fig:SV_flip}E).
Interestingly, the responses are noisier at intermediate contrasts (Fig. \ref{fig:SV_flip}F).
This suggests a new hypothesis explaining phenomena of variability in V1: noise is greater in stabilized supralinear networks during winner-take-all competition between the S- and V- populations, but is quenched as either the S- or V-population overtakes the other.

@@

The dynamics are driven by the rectified and exponentiated sum of recurrent inputs $Wx$, external inputs $h$, and external noise $\epsilon \sim \mathcal{N}(0, \sigma_{\epsilon}^2)$:

\begin{equation}
T \frac{dx}{dt} = -x + x_{\text{rest}} +Wf_r(x) + b + h + \epsilon
\end{equation}

where $T$ is a diagonal matrix with $\tau_E = 20\text{ms}$ and $\tau_P = \tau_S = \tau_V = 10\text{ms}$, $x_{\text{rest}} = -65\text{mV}$, and $f_r(x) = \left[x - x_{\text{rest}} \right]_+^2$.

The noise is an modeled as an Ornstein-Uhlenbeck process:
\begin{equation}
\tau_{\text{noise}} d\epsilon_\alpha = -\epsilon_\alpha dt + \sqrt{2\tau_{\text{noise}}}\sigma_\alpha dB
\end{equation}

where $\tau_{\text{noise}}$ is slow at 50ms, $\sigma_E = 0.2\text{mV}$, $\sigma_P = \sigma_S = \sigma_V = 0.1\text{mV}$, and $dB$ is a standard Wiener process.

We considered fixed effective connectivity weights $W$ approximated from experimental recordings of publicly available datasets of mouse V1 \cite{allen2018layer, billeh2019systematic} (see Section \ref{methods_V1}).  The input to this circuit is comprised of a nominal baseline input $b$ and additive contrast-dependent input $h$
\begin{equation}
h = c \begin{bmatrix} h^{(c)} \\ h^{(c)} \\ 0 \\ 0 \end{bmatrix}.
\end{equation}

@@

%\begin{figure}[h]
%\caption{\small \textbf{A}. ...
% }\label{fig:drdh}
%\begin{center}
%\includegraphics[scale=.3]{figs/FigSabc/FigSabc.pdf}
%\end{center}
%\end{figure}
%
%s
%\begin{figure}[h]
%\caption{\small \textbf{A}. ...
% }\label{fig:drdh}
%\begin{center}
%\includegraphics[scale=.15]{figs/FigSX/FigSX.pdf}
%\end{center}
%\end{figure}

@@

% With this model, we are interested in the differential responses of each neuron-type population to changes in input $dh$. 
% Initially, we studied the linearized response of the system to input $\frac{dx_{ss}}{dh}$ at the steady state response $x_{ss}$, i.e. a fixed point. 
% All analyses of this model consider the steady state response, so we drop the notation $ss$ from here on.
% While this linearization accurately predicts differential responses $dx = \left[ dx_{E} , dx_{P} , dx_{S} ,dx_{V} \right]^\top$  for small differential inputs to each population $dh = \left[ 0.1 , 0.1 , 0.1 , 0.1 \right]^\top$ (Fig \ref{fig:V1_EPI}B left), the linearization is a poor predictor in this nonlinear model more generally (Fig. \ref{fig:V1_EPI}B right).  
% Currently available approaches to deriving the steady state response of the system are limited.
